---
title: "U.S 2020 Election spread analysis"
author: "Pratyush Singh"
date: "27/07/2020"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

This is a report on 2020 U.S election votes spread from summarising data of various pollster's reports as of July2020 . The original data was obtained from [this page](https://projects.fivethirtyeight.com/polls-page/president_polls.csv).

We are going to use the following library:
```{r loading-libs, message=FALSE}
library(tidyverse)
library(dslabs)
library(ggplot2)
library(prettydoc)
```

# DATA PREPARATION


Initial Table of Data .
```{r}
Polls_init <- read.csv("data/president_polls.csv")
```
```{r , echo=FALSE}
 head(Polls_init)
```

Filter for two major candidate (Donald j. Trump & Joesph Biden), High graded Pollsters & 
Selected relevant variables .
Grouped by Unique Question_id for each poll.
```{r}
poll_filter <- Polls_init %>% filter(candidate_name %in% c("Donald Trump","Joseph R. Biden Jr.") & (fte_grade %in% c("A+", "A", "A-", "B+") | is.na(fte_grade))) %>% 
  select("question_id","state","pollster","fte_grade","sample_size","population","start_date","end_date","candidate_name","pct") %>%
  group_by(question_id)
```
```{r , echo=FALSE}
head(poll_filter)
```
Converted Two candidates as variables for comparision in following steps

-Seperated both variables into two different dataframes
- Merged both variable by unique question_id
```{r}
trump2020 <- poll_filter %>% rowid_to_column() %>%spread(candidate_name , pct) %>% select( -rowid , -`Joseph R. Biden Jr.`) %>% filter(!is.na(`Donald Trump`))
joe_biden2020  <- poll_filter %>% rowid_to_column()%>%spread(candidate_name , pct) %>% filter(!is.na(`Joseph R. Biden Jr.`))%>% select( question_id , `Joseph R. Biden Jr.`) 

poll_2020 <- merge(trump2020 , joe_biden2020 , by = "question_id")

```
```{r , echo=FALSE}
head(poll_2020)
```

Added collumn for spread [ $spread = biden/100 - trump/100$] and Saved rda file ;

-Spread specifies the difference between relative probabilities .

```{r}
poll2020 <- poll_2020 %>% mutate(spread = `Joseph R. Biden Jr.`/100 -`Donald Trump`/100 ) 

save(poll2020 , file = "rda/poll2020.rda")
```

```{r , echo=FALSE}
head(poll2020)
```

# EXPLORATORY DATA ANALYSIS 

Loading the tidy data
```{r}
load( "rda/poll2020.rda")
```
The data of various pollster has been summarised to give more accurate result .
This makes for less poll to poll bias and larger sample sizes 


**Initial spread** (without general bias , state- wise bias and poll to poll bias) is ;

$Average = \sum(spread\times samplesize)/\sum(samplesize)$

```{r}
results <- poll2020 %>%   summarise(avg = sum(spread*sample_size)/sum(sample_size))
spread <- round(results$avg*100,1)
```

**SPREAD**

$Spread = round(Average)$
```{r , echo=FALSE}
spread

```
This Spread shows that Joe Biden is Comfortably ahead in polls by <span style="color: red;">**8.1**</span>  percent points 



**STANDARD DEVIATION** 
$sd = standard_deviation(spread)/\sqrt(n)$

```{r}
sd <- sd(poll2020$spread)/sqrt(length(poll2020$spread))
```

```{r , echo=FALSE}
sd
```
 <span style="color: red;">**0.0057956**</span>

**95% CONFIDENCE INTERVAL FOR CANDIDATE JOE BIDEN**
```{r}
tibble(lower_bar =results$avg -qnorm(0.965)*sd , upper_bar=results$avg +qnorm(0.965)*sd) 
```
 There is 95% chance of spread to be between <span style="color: red;">**0.0705911 -	0.09159332**</span>	

**UN-BIASED ANALYSIS** 

Taking account of different biases in polling data 
The statistics are re-analysed using Bayesian Sats .

**BIASES**
```{r}
tau <- 0.035  # general bias in U.S poll results from hitorical data
sigma_b <- 0.03 # median state-wise bias from last year's election
mu <- 0.0 # pollster bias is taken as 0 , assuming for each pollster biased towards Trump there is one                  biased towards Biden
sigma <- sqrt(sd^2 + sigma_b^2)
B <-  sigma^2 /(sigma^2 + tau^2)
# Bayesian variables
```

**UN-BIASED SPREAD**
$Biased Average = B \times \mu + (1 -B) \times(Average)$
```{r}
p_mean <- B*mu + (1-B)*results$avg
spread_biased <- round(p_mean*100 , 1)
```
```{r , echo=FALSE}
spread_biased

```
 This Spread shows that Joe Biden is still ahead in polls even after taking the genral bias in account by  <span style="color: red;">**4.6**</span> percent points

**UN-BIASED STANDARD ERROR**
SE = $\sqrt(1 / (1/(\sigma)^2 + 1/(\tau)^2)$
```{r}
p_se <- sqrt(1 /(1/sigma^2 + 1/tau^2))
```
```{r , echo = FALSE}
p_se

```
 <span style="color: red;">**0.02301765**</span>


**UN-BIASED 95% CONFIDENCE INTERVAL** 
```{r}
tibble(lower_bar =p_mean -qnorm(0.965)*p_se , upper_bar=p_mean +qnorm(0.965)*p_se) 
```
 There is 95% chance of spread to be between <span style="color: red;">**0.004313939 -	0.08772578**</span>	

**PROBABILITY OF JOE BIDEN TO WIN THE POPULAR VOTE**
```{r ,echo=FALSE}
(1 - pnorm(0 , p_mean,p_se))*100
```
There is  <span style="color: red;">**97.72%**</span> probabilty of Joe Biden to win the popular votes. 

**VISUAL COMPARISON OF TWO MODELS **
```{r , echo=FALSE}
spread_data <- tibble(stat = c("Biased" , "Unbiased") , avg = c(results$avg , p_mean) ,
                      se = c(sd , p_se))
```

```{r , echo= FALSE}
spread_data %>% ggplot() + geom_pointrange(aes(stat , avg ,
                                                              ymin = avg - qnorm(0.965)*se ,
                                                              ymax = avg +qnorm(0.965)*se , 
                                                              color = stat )) + theme_dark()
```

## CONCLUSION

Although Unbiased analysis increases the confidence interval significantly , as well as reduces the spread by almost half . The statistics still shows the polls to be significantly in favour of Joe Biden and none of the confidence interval include 0 .
So we can confidently conclude the polls to be in favour of Joe Biden as of July 2020



